Comparing Machine Learning algorithms for classifying cognitive states from fMRI data

### 2012-11-12 article/article.tex
review

%% Abstract section.
\abstract{
Functional magnetic resonance imaging (fMRI) is a brain mapping technique that allows researchers to infer activity in a subject's brain. This technique is typically used to determine which areas of the brain are involved in performing a certain task. Using this knowledge, we may also be able to determine what task was performed by looking at fMRI data. In this paper, we present a comparison of the performance of machine learning algorithms in between-subjects classification of cognitive states, using fMRI data. We found that averaging across ROIs gave the best classification performance. All classifiers that we used performed above-chance, and we found that a Multivariate Gaussian classifier outperformed a Naive Bayesian classifier and a Support Vector Machine. Overall we can say that machine learning algorithms provide a feasible method for decoding cognitive states from fMRI data.
} % end of abstract

@@

Traditionally, data from each voxel, or from a group of voxels, are analysed with analysis of variance. Voxels are usually convoluted with a haemodynamic response function in an attempt to infer the spiking neural activity from the slowly building and dropping BOLD response \cite{he:2002na} (see Figure~\ref{fig:bold_response}). In this setup, experimental conditions are the independent variables and a general linear model is fitted to each voxel, or group of voxels. By using this technique, voxels or brain areas can be identified that differ in their average BOLD response across conditions.\\
\indent Recent studies indicate that this approach may have some serious limitations. One study used a dataset of 1326 subjects, which is much higher than the usual 15 to 25 subjects found in fMRI studies \cite{thy:2012very}. This study explored what happened to a simple contrast between BOLD signal while watching faces and baseline activity, when increasing the number of subjects included in the analysis. They found that, as the number of analysed subjects increased, an increasing number of brain areas that extended well beyond the expected visual- and face recognition areas showed significant deviation from baseline. Another study, using only three subjects, repeated a simple letter/number discrimination task and visual stimulation. The notable part is that this very simple experiment was run 100 times for each subject, again resulting in exceptionally high signal-to-noise ratio compared to conventional fMRI experiments. In this study the authors again found widespread significant BOLD-responses \cite{go:2012whole}. These findings suggest that significant deviations from baseline activity do not necessarily indicate any special involvement in the performed task, because almost the entire brain will show significant deviations given a high enough signal-to-noise ratio. These findings are perhaps not surprising, because everything we know about the brain indicates that it is best seen as a network with countless connections to other parts of the brain. Therefore, it makes sense to also analyse signals from the brain as a pattern of information coming from the entire brain, rather than attempting to link specific brain areas to experimental conditions.\\
\indent In fact, Machine learning algorithms have been employed in studies to learn the patterns of activity associated with some experimental condition. Subsequently, these learned patterns can be used on unseen subjects in order to predict experimental conditions from their brain activity patterns. If these predictions are accurate then it means that the identified patterns hold information about the brain activity associated with some experimental task.\\
\indent The general goal of these machine learning studies is to identify what experimental condition a subject is in at a certain time. The state that a person is in during some trial (e.g., looking at a face) is referred to as \emph{cognitive state}, because the cognitive processes that are associated with being in that state are of interest. Specifically, the aforementioned studies try to infer the cognitive state of a subject from the brain activity that is evoked in a subject by performing that trial. For example, in an experiment where subjects look at a set of pictures (e.g., a star and a plus), they would be in a certain cognitive state that is the result of looking at these pictures. One could then attempt to use machine learning on the fMRI signal that is evoked by the the brain activity associated with these cognitive states, and use this to learn what patterns of activity in the brain are associated with that cognitive state. This \emph{StarPlus} experiment has been performed by researchers where, in addition to viewing pictures, subjects also were tasked to read sentences containing some true or false statement about the pictures. These data have then been used to successfully infer the cognitive states of looking at either sentences or pictures within single subjects \cite{mi:2003within}, and between subjects \cite{wa:2003betw}.\\
\indent In this study we use the same StarPlus data as in the aforementioned experiment, and use machine learning algorithms to decode the cognitive states of subjects while looking at pictures or while looking at sentences. An interesting difference between many applications of machine learning, and this study is that most machine learning algorithms have the goal of achieving the maximum possible discrimination between conditions or classes. However, for fMRI studies it is important that the pattern of activity that is identified by the algorithm is a pattern associated with the cognitive state. When using non-linear classifiers, it is possible to decode features predictive of the cognitive state, where the brain does not actually respond to these features. For example, when decoding the primary visual cortex with high enough spatial resolution, it would be possible to combine simple features with a non-linear neural network to decode the presence or absence of, for example, faces. However, this information is not actually processed in that part of the brain \cite{to:2012rev}. To avoid this problem it is recommended to only use linear classifiers when decoding cognitive states from fMRI data. The classifiers that we compare in this paper are a Gaussian Naive Bayesian (GNB) classifier, Linear Support Vector Machine (LSVM), and a Multivariate Gaussian (MVG) classifier.\\
\begin{figure}
	\centering
	\includegraphics[width=60mm]{figures/f1_bold_response}
  	\caption{Above is the hypothetical neural response to some stimulus. Below is the BOLD response that is evoked by the short spike of neural activity. The neural activity shown above is attempted to be inferred from the BOLD response below that is measured by the fMRI scanner. Figure is adopted from Heeger and Ress \cite{he:2002na}.}
  	\label{fig:bold_response}
\end{figure}
\indent The goal of this study is to compare the classification performance of the cognitive states associated with looking at a picture or looking at a sentence between subjects. This objective introduces an important issue, namely, how the data from the fMRI scans can be compared between subjects. There are two main ways this could be achieved. First, groups of voxels that are anatomically defined as belonging to a certain brain region, a so-called region of interest (ROI), can be averaged into a single \emph{supervoxel}, which can be compared between subjects. Second, a talairach coordinate system can be used that maps the voxels of each subject onto an anatomically defined coordinate. The ROI averaging it is a rather crude method, and a lot of information of the individual voxels is lost. However, because the ROIs are relatively easy to identify by an expert they can be compared easily between subjects. By using the Talairach coordinate system the data of each individual voxel can be used~\cite{talairach1988co}. However, because the error of doing this is usually in the order of a centimetre, while the size of the voxels are in the order of millimetres, the coordinates are difficult to generalise. Because we aim to do between-subjects comparisons the generalisation performance is important, thus we will use averaged ROIs in order to compare the classifiers \cite{mi:2004coord}.

>>>>

%% Abstract section.
\abstract{
Functional magnetic resonance imaging (fMRI) is a brain mapping technique that allows researchers to infer activity in a subject's brain. This technique is typically used to determine which areas of the brain are involved in performing a certain task. Using this knowledge, we may also be able to determine what task was performed by looking at fMRI data. In this paper, we present a comparison of the performance of machine learning algorithms in between-subjects classification of cognitive states, using fMRI data. We found that averaging across Regions of Interest (ROIs) gave the best classification performance. All classifiers that we used performed above-chance, and we found that a Multivariate Gaussian classifier outperformed a Naive Bayesian classifier and a Support Vector Machine. Overall we can say that machine learning algorithms provide a feasible method for decoding cognitive states from fMRI data.
} % end of abstract

@@

Traditionally, data from each voxel, or from a group of voxels, are analyzed with analysis of variance. Voxels are usually convoluted with a haemodynamic response function in an attempt to infer the spiking neural activity from the slowly building and dropping BOLD response \cite{he:2002na} (cf.~Figure~\ref{fig:bold_response}). In this setup, experimental conditions are the independent variables and a general linear model is fitted to each voxel, or group of voxels. By using this technique, voxels or brain areas can be identified that differ in their average BOLD response across conditions.

Recent studies indicate that this approach may have some serious limitations. One study used a dataset of 1326 subjects, which is much higher than the usual 15 to 25 subjects found in fMRI studies \cite{thy:2012very}. This study explored what happened to a simple contrast between BOLD signal while watching faces and baseline activity, when increasing the number of subjects included in the analysis. They found that, as the number of analyzed subjects increased, an increasing number of brain areas that extended well beyond the expected visual- and face recognition areas showed significant deviation from baseline. Another study, using only three subjects, repeated a simple letter/number discrimination task and visual stimulation. The notable part is that this very simple experiment was run 100 times for each subject, again resulting in exceptionally high signal-to-noise ratio compared to conventional fMRI experiments. In this study the authors again found widespread significant BOLD-responses \cite{go:2012whole}. These findings suggest that significant deviations from baseline activity do not necessarily indicate any special involvement in the performed task, because almost the entire brain will show significant deviations given a high enough signal-to-noise ratio. These findings are perhaps not surprising, because everything we know about the brain indicates that it is best seen as a network with countless connections to other parts of the brain. Therefore, it makes sense to also analyze signals from the brain as a pattern of information coming from the entire brain, rather than attempting to link specific brain areas to experimental conditions.

In fact, Machine learning algorithms have been employed in studies to learn the patterns of activity associated with some experimental condition. Subsequently, these learned patterns can be used on unseen subjects in order to predict experimental conditions from their brain activity patterns. If these predictions are accurate then it means that the identified patterns hold information about the brain activity associated with some experimental task.

The general goal of these machine learning studies is to identify what experimental condition a subject is in at a certain time. The state that a person is in during some trial (e.g., looking at a face) is referred to as \emph{cognitive state}, because the cognitive processes that are associated with being in that state are of interest. Specifically, the aforementioned studies try to infer the cognitive state of a subject from the brain activity that is evoked in a subject by performing that trial. For example, in an experiment where subjects look at a set of pictures (e.g., a star and a plus), they would be in a certain cognitive state that is the result of looking at these pictures. One could then attempt to use machine learning on the fMRI signal that is evoked by the the brain activity associated with these cognitive states, and use this to learn what patterns of activity in the brain are associated with that cognitive state. This \emph{StarPlus} experiment has been performed by researchers where, in addition to viewing pictures, subjects also were tasked to read sentences containing some true or false statement about the pictures. These data have then been used to successfully infer the cognitive states of looking at either sentences or pictures within single subjects \cite{mi:2003within}, and between subjects \cite{wa:2003betw}.

In this study we use the same StarPlus data as in the aforementioned experiment, and use machine learning algorithms to decode the cognitive states of subjects while looking at pictures or while looking at sentences. An interesting difference between many applications of machine learning, and this study is that most machine learning algorithms have the goal of achieving the maximum possible discrimination between conditions or classes. However, for fMRI studies it is important that the pattern of activity that is identified by the algorithm is a pattern associated with the cognitive state. When using non-linear classifiers, it is possible to decode features predictive of the cognitive state, where the brain does not actually respond to these features. For example, when decoding the primary visual cortex with high enough spatial resolution, it would be possible to combine simple features with a non-linear neural network to decode the presence or absence of, for example, faces. However, this information is not actually processed in that part of the brain \cite{to:2012rev}. To avoid this problem it is recommended to only use linear classifiers when decoding cognitive states from fMRI data. The classifiers that we compare in this paper are a Gaussian Naive Bayesian (GNB) classifier, Linear Support Vector Machine (LSVM), and a Multivariate Gaussian (MVG) classifier.

\begin{figure}
	\centering
	\includegraphics[width=60mm]{figures/f1_bold_response}
  	\caption{Above is the hypothetical neural response to some stimulus. Below is the BOLD response that is evoked by the short spike of neural activity. The neural activity shown above is attempted to be inferred from the BOLD response below that is measured by the fMRI scanner. Figure is adopted from Heeger and Ress \cite{he:2002na}.}
  	\label{fig:bold_response}
\end{figure}

The goal of this study is to compare the classification performance of the cognitive states associated with looking at a picture or looking at a sentence between subjects. This objective introduces an important issue, namely, how the data from the fMRI scans can be compared between subjects. There are two main ways this could be achieved. First, groups of voxels that are anatomically defined as belonging to a certain brain region, a so-called region of interest (ROI), can be averaged into a single \emph{supervoxel}, which can be compared between subjects. Second, a talairach coordinate system can be used that maps the voxels of each subject onto an anatomically defined coordinate. The ROI averaging it is a rather crude method, and a lot of information of the individual voxels is lost. However, because the ROIs are relatively easy to identify by an expert they can be compared easily between subjects. By using the Talairach coordinate system the data of each individual voxel can be used~\cite{talairach1988co}. However, because the error of doing this is usually in the order of a centimetre, while the size of the voxels are in the order of millimetres, the coordinates are difficult to generalize. Because we aim to do between-subjects comparisons the generalization performance is important, thus we will use averaged ROIs in order to compare the classifiers \cite{mi:2004coord}.
